The Galactic Midplane is Not a Plane: Implications for dynamical analysis with {\em Gaia} data and beyond

### 2019-04-22 paper/ms.tex
Fix line indenting throughout

Underlying the quantitative analysis of many of the mechanisms at work that
give rise to these signatures is the assumption of a global, axisymmetric
Galactocentric coordinate system \citep{2008gady.book.....B} in which the
Sun's position (and therefore the origin of the coordinate system) can be
defined and measured both precisely and accurately. In this system, the center
of the Galaxy is usually taken to be the location of the central supermassive
black hole, Sgr~A\textsuperscript{*} \citep[e.g.][]{2004ApJ...616..872R}. From
stellar motions near Sgr~A\textsuperscript{*}, the distance from the Sun to
the Galactic center
    can be measured
\citep{2009ApJ...692.1075G,2018AA...615L..15G}, with a recent 
    measurement using near-infrared interferometry of $8.178 \pm
    0.035$~kpc 
\citep{2019arXiv190405721A}. The Sun's vertical distance
from the Galactic midplane, the other independent measurement that sets this
coordinate system, can be determined by identifying where the stellar density
and velocities reach a maximum (effectively the median height of all disk
stars), and is usually taken to be $\sim 25\,\pc$ \citep{2001ApJ...553..184C},
with more recent measurements from {\em Gaia} DR2 placing it at $\sim 20\,\pc$
\citep{2019MNRAS.482.1417B}. Another strategy is to use the cold gas or HII
regions in the disk to define the Galactic midplane, leading to slightly
different values ($\sim 5\,\pc$) for the Sun's relative height
\citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em Gaia} review of these
measurements is given by \citet{2016ARAA..54..529B}.

@@

If the system being considered departs from axisymmetry in a significant
and/or non-adiabatic way, the actions computed using an axisymmetric
approximation to the true potential can exhibit cyclic dependence on the
orbital phase (or time at which the star's position and velocity are
observed), large-scale migration, or diffusion from their initial values. In
the Milky Way, stellar actions are expected to diffuse on short time scales
due to scattering with gas clouds and to evolve on longer time scales in the
case of orbits near resonances with spiral arms, bar(s), and other large scale
perturbations \citep{2014RvMP...86....1S}. For this reason, we and other
authors have used actions to study stellar scattering in the Milky Way disk
using the improved astrometry of {\em Gaia} DR2 and various age catalogues
\citep{2018ApJ...867...31B,2018arXiv180803278T}. Actions have also been used
to study different models of spiral structure in the Milky Way
\citep{2019MNRAS.tmp..155S}. Characteristics of the distribution of stars in
the extended solar neighborhood in action space are discussed in
\citet{2019MNRAS.484.3291T}.

@@

In this work we discuss how the imposition of a global axisymmetric coordinate
system based on a local measurement of the midplane location and
Galactocentric distance introduces systematic error in the computation of
actions, especially for stars in the thin disk. Our main point is that the
``local'' Galactic midplane is not the same as the ``global'' Galactic
midplane. For stars relatively far from the Sun, the local midplane location
will not agree with our local midplane extrapolated onto their position,
thanks to relatively small-scale variations in the density of the gas and
stellar distribution. Converting positions and velocities of more distant
stars from a heliocentric to a Galactocentric coordinate system introduces a
systematic bias in the $z$ coordinate, and hence the actions computed for
these stars. The further from the solar neighborhood the target star, the more
likely the mismatch will result in large systematic uncertainty, especially in
the vertical action $J_z$. Depending on the type of orbit being studied, this
sets a limit on the ``dynamical solar neighborhood'' over which the local
coordinate frame can be usefully extended for action-space analysis. A similar
argument applies to 
%mm [given that this is down to 0.3% error, seems a bit academic]
   any remaining 
uncertainty in measurements of 
     the Galactocentric radius of the Sun $R_{\odot}$
and to variations in the local standard of rest (LSR).

In Section~\ref{sec:ref_frame}, we describe the general impact coordinate
system errors have on the measured actions. Lacking an empirical, nonlocal
definition of the global Galactic midplane or a measurement of the expected
midplane variation in the Milky Way itself,  in Section~\ref{sec:local_fire}
we 
    examine the expected 
azimuthal dependence of the local midplane, and
hence the expected size of the dynamical solar neighborhood, using
cosmological, hydrodynamical, zoom-in simulations of Milky Way-mass galaxies
from the Feedback in Realistic Environments (FIRE)
collaboration\footnote{\url{https://fire.northwestern.edu}}
\citep{2016ApJ...827L..23W,2017arXiv170206148H}.

>>>>

Underlying the quantitative analysis of many of the mechanisms at work that
give rise to these signatures is the assumption of a global, axisymmetric
Galactocentric coordinate system \citep{2008gady.book.....B} in which the
Sun's position (and therefore the origin of the coordinate system) can be
defined and measured both precisely and accurately. In this system, the center
of the Galaxy is usually taken to be the location of the central supermassive
black hole, Sgr~A\textsuperscript{*} \citep[e.g.][]{2004ApJ...616..872R}. From
stellar motions near Sgr~A\textsuperscript{*}, the distance from the Sun to
the Galactic center can be measured \citep{2009ApJ...692.1075G,
2018AA...615L..15G}, with a recent measurement using near-infrared
interferometry of $8.178 \pm 0.035$~kpc \citep{2019arXiv190405721A}. The Sun's
vertical distance from the Galactic midplane, the other independent
measurement that sets this coordinate system, can be determined by identifying
where the stellar density and velocities reach a maximum (effectively the
median height of all disk stars), and is usually taken to be $\sim 25\,\pc$
\citep{2001ApJ...553..184C}, with more recent measurements from {\em Gaia} DR2
placing it at $\sim 20\,\pc$ \citep{2019MNRAS.482.1417B}. Another strategy is
to use the cold gas or HII regions in the disk to define the Galactic
midplane, leading to slightly different values ($\sim 5\,\pc$) for the Sun's
relative height \citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em Gaia} review
of these measurements is given by \citet{2016ARAA..54..529B}.

@@

If the system being considered departs from axisymmetry in a significant
and/or non-adiabatic way, the actions computed using an axisymmetric
approximation to the true potential can exhibit cyclic dependence on the
orbital phase (or time at which the star's position and velocity are
observed), large-scale migration, or diffusion from their initial values. In
the Milky Way, stellar actions are expected to diffuse on short time scales
due to scattering with gas clouds and to evolve on longer time scales in the
case of orbits near resonances with spiral arms, bar(s), and other large scale
perturbations \citep{2014RvMP...86....1S}. For this reason, we and other
authors have used actions to study stellar scattering in the Milky Way disk
using the improved astrometry of {\em Gaia} DR2 and various age catalogues
\citep{2018ApJ...867...31B, 2018arXiv180803278T}. Actions have also been used
to study different models of spiral structure in the Milky Way
\citep{2019MNRAS.tmp..155S}. Characteristics of the distribution of stars in
the extended solar neighborhood in action space are discussed in
\citet{2019MNRAS.484.3291T}.

@@

In this work we discuss how the imposition of a global axisymmetric coordinate
system based on a local measurement of the midplane location and
Galactocentric distance introduces systematic error in the computation of
actions, especially for stars in the thin disk. Our main point is that the
``local'' Galactic midplane is not the same as the ``global'' Galactic
midplane. For stars relatively far from the Sun, the local midplane location
will not agree with our local midplane extrapolated onto their position,
thanks to relatively small-scale variations in the density of the gas and
stellar distribution. Converting positions and velocities of more distant
stars from a heliocentric to a Galactocentric coordinate system introduces a
systematic bias in the $z$ coordinate, and hence the actions computed for
these stars. The further from the solar neighborhood the target star, the more
likely the mismatch will result in large systematic uncertainty, especially in
the vertical action $J_z$. Depending on the type of orbit being studied, this
sets a limit on the ``dynamical solar neighborhood'' over which the local
coordinate frame can be usefully extended for action-space analysis. A similar
argument applies to any remaining uncertainty in measurements of the
Galactocentric radius of the Sun $R_{\odot}$ and to variations in the local
standard of rest (LSR).

In Section~\ref{sec:ref_frame}, we describe the general impact coordinate
system errors have on the measured actions. Lacking an empirical, nonlocal
definition of the global Galactic midplane or a measurement of the expected
midplane variation in the Milky Way itself,  in Section~\ref{sec:local_fire}
we examine the expected azimuthal dependence of the local midplane, and hence
the expected size of the dynamical solar neighborhood, using cosmological,
hydrodynamical, zoom-in simulations of Milky Way-mass galaxies from the
Feedback in Realistic Environments (FIRE)
collaboration\footnote{\url{https://fire.northwestern.edu}}
\citep{2016ApJ...827L..23W, 2017arXiv170206148H}.
